Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-03-21 article.tex
Rewrite introduction to clarify definition, translation and solving

\Acp{CRN} are a fundamental concept of modelling
in numerous fields including systems biology, biochemistry, chemical
engineering and epidemiology. They are comprised of a set of chemical
species or biological entities
and a set of reactions that mediate transformations between them. These
systems can be modelled and simulated through multiple approaches:
deterministic \acp{ODE} to model
macroscopic behavior, \acp{SDE} to model
microscopic fluctuations, and jump processes (Gillespie-like
simulations) to account for the discreteness of populations. Instead of
directly writing the equations for each of these formulations, which is
error-prone and difficult to reuse, these models can be written in a
higher-level description that can be compiled for these different types
of simulations.

Several tools already exist for defining and simulating \acp{CRN}.
BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022}, a registry of simulation tools,
list at least 15 Python software including COPASI \cite{hoopsCOPASICOmplexPAthway2006},
Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and PySB \cite{lopezProgrammingBiologicalModels2013}. COPASI is a
standalone software with a \ac{GUI} for defining and simulating models. It's
widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow
advanced scripting. Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to
simulate models. PySB is a Python library that created a \ac{DSL} using
standard Python to define models, which are then compiled to \acp{ODE} using a
Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.

One limitation of these tools is that they are not extensible from
Python, as they are not fully Python packages but wrap libraries in
other languages to do the simulation. While they enable model definition
and simulation control via Python scripts, they don't fully leverage
Python's extensive package ecosystem. For example, in COPASI, users are
restricted to predefined distributions for Random Parameter Scans and
cannot utilize the diverse distributions available in
\texttt{scipy.stats}. Similarly, Tellurium doesn't allow the use of
Python solvers, as adding new integrators requires C++ implementation.

@@

Another challenge is the way models are written. Many tools use a \ac{DSL} for coding, and support the
\ac{SBML} as an exchange format \cite{huckaSBMLL3V2}, as
direct \ac{SBML} coding is impractical. A \ac{DSL} allows reuse of code in
different programming environments, but it will not be recognized by
default in \acp{IDE}. Therefore \acp{DSL}
cannot provide the development experience such as syntax highlighting,
code completion, refactoring, and static analysis, unless such support
is specifically developed. Tellurium uses a
\ac{DSL} called Antimony \cite{smithAntimonyModularModel2009}, which can be translated to
\ac{SBML}. An extension for Visual Studio Code was developed, but its
maintenance could be a demanding task for the systems biology community.
PySB, using Python's dynamic nature, created a \ac{DSL} within Python. By
default, it uses global state to create species and parameters, without
assigning them to Python variables or adding them explicitly to the
model, but this approach is not fully compatible with \acp{IDE}, affecting
the development experience.

To overcome these limitations, we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows one to define differential equation systems, while SimBio
builds on it for defining reaction networks. They are focused on
providing an ergonomic experience to end-users by integrating well with
\acp{IDE} and static analysis tools through the use of standard modern Python
syntax. Since they are coded in Python, every part from model definition
to simulation can be extended from Python syntax. Being the first-ever
pure Python packages for systems modelling, they offer extensive
extensibility, from simple tasks like reusing integrators defined in
other packages, to complex ones like altering the compilation process to
leverage some structure in the equations. For example, using a for-loop
in the compiled equations could improve the runtime performance if there
is some repetitive structure in the system, as happens in spatial
modelling. The models built using these packages can be introspected to
create other representations, such as graphs connecting species and/or
reactions, or tables with parameters or equations. Furthermore, they
have a modular architecture with a clear separation of concerns, making
it easier to maintain or to contribute new code, which is beneficial for
developers and maintainers. We showcased the reliability of these tools
by benchmarking them against the simulation results from other tools. We
also highlighted the substantial performance improvements our tools
offer, as this is crucial for construction and simulation of models of
whole cells and organisms, which necessitate the simulation of
significantly large-scale models.

>>>>

\Acp{CRN} are a fundamental concept of modelling in numerous fields
including systems biology, biochemistry, chemical engineering and epidemiology.
They are comprised of a set of chemical species or biological entities
and a set of reactions that mediate transformations between them.
These systems can be simulated through multiple approaches:
deterministic \acp{ODE} to model macroscopic behavior,
\acp{SDE} to model microscopic fluctuations,
and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
Instead of directly writing the equations for each of these formulations,
which is error-prone and difficult to reuse,
these models can be \emph{defined} in a higher-level description
that can be \emph{translated} into equations for the different types of simulations
and, then, \emph{solved} numerically.

Several tools already exist to \emph{define}, \emph{translate} and \emph{solve} \acp{CRN}.
BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022},
a registry of simulation tools,
lists at least 15 Python software including
COPASI \cite{hoopsCOPASICOmplexPAthway2006},
Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and
PySB \cite{lopezProgrammingBiologicalModels2013}.
COPASI is a standalone software with a \ac{GUI}
that is widely used for its user-friendly interface and comprehensive features.
It also includes Python bindings \cite{bergmannBASICOSimplifiedPython2023b} that allow advanced scripting.
Tellurium is a Python-based modeling environment
that uses a C++ library called libRoadRunner in the backend to translate and solve models.
PySB is a Python library that created a \ac{DSL} using standard Python to define models,
which are then translated to \acp{ODE} using a Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.

One limitation of these tools is their extensibility from Python.
As they wrap libraries in other languages for defining, translating and/or solving models,
these steps cannot be altered or easily inspected from Python.
While they enable model definition and running simulations via Python scripts,
they cannot fully leverage Python's extensive package ecosystem.
For example,
COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
and adding new integrators requires working with C++.
In particular, the step that translates into equations is not exposed by any of these tools.
As such, it is not possible to apply custom optimizations to the equations
or use automatic differentation packages such as JAX \cite{jax2018github} to compute the model's jacobian.

@@

Another challenge is the way models are defined.
Many tools support the \ac{SBML} \cite{huckaSBMLL3V2} as an exchange format,
a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
As writing \ac{SBML} directly is impractical,
Tellurium uses a \ac{DSL} called Antimony \cite{smithAntimonyModularModel2009} for defining models.
\Acp{DSL} allows to reuse the same code in different programming environments,
but are not recognized by default in \acp{IDE}
and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
For Antimony, an extension providing this capabilities was developed for Visual Studio Code,
but its maintenance could be a demanding task for the systems biology community.
In the case of PySB,
using Python's dynamic nature,
it developers designed a \ac{DSL} within Python.
By default, it uses the global scope to create species and parameters,
without explicitly assigning them to Python variables or to the model,
but this approach is not fully compatible with \acp{IDE},
affecting the development experience.

To overcome these limitations,
we developed poincaré and SimBio,
open-source Python packages for defining, translating and solving systems.
Poincaré allows one to define differential equation systems
using variables, parameters and constants,
and assiging rate equations to variables.
For defining \acp{CRN},
SimBio builds on top of poincaré providing species and reactions that keep track of stoichiometries.
Both are focused on providing an ergonomic experience to end-users
by integrating well with \acp{IDE} and static analysis tools
through the use of standard modern Python syntax.
Moreover, since they are coded in pure Python,
each step from model definition, translation to equations or solving
can be extended or debugged from Python.
Being the first-ever pure Python packages for systems modelling,
they offer extensive extensibility,
from simple tasks like reusing integrators defined in other packages,
to complex ones like altering the compilation process to leverage some structure in the equations.
For example, using a for-loop in the compiled equations could improve the runtime performance
if there is some repetitive structure in the system,
as happens in spatial modelling.
The models built using these packages can be introspected to create other representations,
such as graphs connecting species and/or reactions, or tables with parameters or equations.
Furthermore, they have a modular architecture with a clear separation of concerns,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.
We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
We also highlighted the substantial performance improvements our tools offer,
as this is crucial for construction and simulation of models of whole cells and organisms,
which necessitate the simulation of significantly large-scale models.
